Epistasis is widespread in the control of human gene expression

### 2013-05-28 docs/manuscript/nature_letter.tex
writing continues

\title{Epistasis identified to be widespread in the control of human gene expression}

@@

\begin{abstract}
A long standing question in human genetics is the extent to which epistasis, where one mutation's effect on a trait depends on other mutations present in the genome, contributes to complex traits. Though epistasis is known to arise from artificial gene manipulation studies in model organisms, and examples have been shown in other species, few convincing examples exist for epistasis amongst natural polymorphisms in human traits. Its absence from empirical findings could be due to its unimportance in the genetic control of complex traits, or it may be too technically difficult to detect due to statistical power and computational issues. Here we show that, using advanced computation techniques and a gene expression study design where any effects are expected to be large, abundant evidence for epistasis is found. In a modest sample size of 842 individuals with data on 7339 gene expression levels in whole blood, we find that after correcting for multiple testing the expression of 249 genes is controlled by 549 significant pairwise epistatic interactions. We attempted replication in two independent datasets and 421 show evidence of significance in at least one dataset. Several genes are controlled by epistatic interactions whereby one cis-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs, for example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by trans-SNPs on 18 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each cis-trans interaction. This study presents the first strong evidence for the widespread existence of epistasis among natural polymorphisms in humans.
\end{abstract}

@@

The past decade has seen a tremendous amount of activity in mapping genetic polymorphisms that underlie complex traits. Typically, SNPs are treated as contributing linearly, independently, and cumulatively to the mean of a trait and this has been successful in identifying hundreds of causal variants. Yet outside the prism of association studies there is widespread evidence for epistasis, for example at the molecular level in artificial double gene knockouts in experimental organisms and at the evolutionary level in fitness adaptation and speciation. Though its importance is frequently the subject of debate there is little convincing empirical evidence for epistasis playing a substantial role in the architecture of complex traits.

There are two major obstacles in detecting epistasis. First, the detection of epistasis is significantly hampered by power issues. For example, there is a greater dependency on linkage disequilibrium (LD) between causal variants and observed SNPs in order to capture non-additive genetic variance, and this problem is exacerbated further with epistasis because there must be high LD at all interacting loci. Relatedly, the dependence on a much larger sample size in order to detect similar effect sizes compared to additive effects, because of the increased number of parameters in statistical tests for interaction terms, and the increased multiple testing correction for testing combinations of SNPs.  Second, one must extend genome-wide association scans into two or more dimensions in order to detect interacting SNPs, resulting in an overwhelming computational burden. We overcome this problem by distributing the search grid across graphics processing unit compute clusters.

>>>>

\title{Epistasis is widespread in the control of human gene expression}

@@

\begin{abstract}
A long standing question in human genetics is the extent to which epistasis, where one mutation's effect on a trait depends on other mutations present in the genome, contributes to complex traits. Though epistasis is known to arise from artificial gene manipulation studies in model organisms, and examples have been shown in other species, few convincing examples exist for epistasis amongst natural polymorphisms in human traits. Its absence from empirical findings could be due to its unimportance in the genetic control of complex traits, or it may be too technically difficult to detect due to statistical power and computational issues. Here we show that, using advanced computation techniques and a gene expression study design where any effects are expected to be large, abundant evidence for epistasis is found. In a modest sample size of 842 individuals with data on 7339 gene expression levels in whole blood, we found that after correcting for multiple testing the expression of 249 genes is controlled by 549 significant pairwise epistatic interactions. We attempted replication in two independent datasets and 421 show evidence of significance in at least one dataset. Several genes are controlled by multi-locus epistatic interactions whereby one cis-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs, for example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by trans-SNPs on 18 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each cis-trans interaction. This study presents the first strong evidence for the widespread existence of epistasis among natural polymorphisms in humans.
\end{abstract}

@@

The past decade has seen a tremendous amount of activity in mapping genetic polymorphisms that underlie complex traits. Typically, SNPs are treated as contributing linearly, independently, and cumulatively to the mean of a trait and this has been successful in identifying hundreds of causal variants. Yet outside the prism of association studies there is widespread evidence for epistasis, not only at the molecular scale in artificial double gene knockouts in experimental organisms but also at the evolutionary scale in fitness adaptation and speciation. Though its importance is frequently the subject of debate so far there is little convincing empirical evidence for epistasis playing a substantial role in the architecture of complex traits.

The detection of epistasis is hampered by power issues for several reasons, including increased dependence on linkage disequilibrium between causal SNPs and observed SNPs, increased model complexity in fitting interaction terms, and more extreme significance thresholds to account for increased multiple testing. When genetic effect sizes are small, as is expected in most complex traits of interest, the power to detect epistasis diminishes rapidly. To overcome this problem we searched for epistasis controlling gene expression levels. Though they number in the thousands, these traits are typically heritable but much less polygenic, thus it is expected that any genetic effects will be relatively large, maximising the chance at detecting epistasis if it should exist.

We searched for pairwise epistasis exhaustively in the BSGS dataset which comprises 842 individuals who are genotyped at 528,000 SNPs and who have gene expression levels measured in whole blood samples for 7339 genes. Recent hardware and software advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis. We used permutation analysis to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ for $p$-values at the 5\% family-wise error rate (FWER).
